---
title: NYPD Shooting Incident Data Report
author: Trey Hendrix
date: |
  Date Created: 260307
  Date Modified: 260323
output: 
  html_document:
    toc: true
    number_sections: true
    
---

```{r Setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE, 
  warning = FALSE
)
```

# Introduction

In this report, I analyze [the historic NYPD shooting incident dataset from data.gov](https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic). I focus on the demographics of shooters and victims and explore the implications of missing perpetrator information. 

```{r Packages}
library(tidyverse)
library(kableExtra) # for more aesthetic and customizable data tables
library(jsonlite) # for parsing JSON metadata 
library(naniar) # for visualizing and analyzing missing data
library(scales) # for graph labels
```

```{r R version check, include=FALSE}
# Check for R version 4.1.0 or higher for the base pipe |>
if (getRversion() < "4.1.0") {
  stop(paste(
    "Error: R version 4.1.0 or higher is required to use the base pipe (|>).",
    "Your current version is", getRversion(),
    ". To run this report, please update R at https://cran.r-project.org/"
  ))
}
```

# Importing Data and Metadata

Using the ["ARCHIVED_NYPD Shooting Incident Data (Historic)" on the data.gov website](https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic), we can copy the link for downloading the .csv format of the dataset. Similarly, we can also grab the metadata via this page (which is a JSON).
```{r Import data}
raw_df <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
meta_list <- fromJSON("https://catalog.data.gov/harvest/object/e150b92a-5c72-4adc-8df5-f27a0a0b641b")
```

```{r Data overview}
glimpse(raw_df)
```
Most of these columns are self-explanatory; however, we will need to dig into the metadata to make sure we understand a few of them. 

```{r Column Definitions}
column_meta_list <- meta_list$distribution |> 
  filter(mediaType == "application/json") |> 
  pull(downloadURL) |> 
  fromJSON()

column_def_df <- column_meta_list$meta$view$columns

column_def_df |> 
  filter(fieldName %in% str_to_lower(c("X_COORD_CD", "Y_COORD_CD"))) |> 
  select(fieldName, description)
```

For our geographic data, we will stick to the standard latitude and longitude fields, so I will drop these variables in the section below.

# Data Cleaning

First, I will also perform some quick cleaning, such as consistently coding missing values and renaming columns. 

```{r Variable selection and name cleanup}
null_str_values <- c("(null)", "UNKNOWN") # We will replace these with NAs in R to better identify and deal with missing data

df <- raw_df |> 
  rename_with(.fn = str_to_lower) |>  # make the column names lowercase for consistency
  mutate(
    occur_date = mdy(occur_date), # mdy = month day year date format 
    across( # clean up NA values stored as strings 
      where(is.character), 
      .fn = function(x) {ifelse(x %in% null_str_values, NA_character_, x)}
      ), 
    precinct = as.character(precinct) # These are not meaningful as integers (they are categories)
  ) |> 
  select(-c(x_coord_cd, y_coord_cd)) # drop unwanted columns

```

## Missing Data

From the previous `glimpse` call above, it is clear that many records have missing values. The `naniar` R package is my favorite way to understand patterns in missing data. First, let's see which variables have missing records:

```{r Missing counts}
gg_miss_var(df)
df |> 
  miss_var_summary() |> 
  mutate(pct_miss = str_glue("{as.numeric(round(pct_miss, 1))}%")) |> # drop the odd S3 data type
  kable() |> 
  kable_classic_2(full_width = FALSE)
```

Fortunately, the majority of records only have a small portion of missing records for most variables (`longitude` to `boro` in the graph above). These variables are also fairly essential/fundamental to understanding a crime that I feel comfortable dropping any records that are missing these information. However, a handful of variables are frequently missing. The location information `loc_classfctn_desc` and `loc_of_occur_desc` are missing too frequently to be informative, so I will largely ignore them. `location_desc` is available for 40% of records while information about the perpetrator (`perp_age_group`, `perp_race`, and `perp_sex`) are available for the majority of incidents. I am interested in digging into this pattern of missing data and will do so in [the second analysis section](# Analysis 2 - The Hidden World of Perps), so I will not drop or impute these data.


**Bias Observation #1**: Approximately half of our records are missing information about the race and sex of the perpetrator, while the victims' race and sex is never missing. We need to keep this bias in the data in mind when analyzing perpetrator-victim dynamics. This bias is somewhat intuitive: crimes are reported by victims, not perpetrators.

```{r Critical NA drop}
# Dropping records with critical missing information
df <- df |> 
  drop_na(jurisdiction_code, vic_age_group, vic_race, lon_lat, longitude, latitude)

# Creating a column to track when information is missing about a perp
df <- df |> 
  mutate(missing_perp_info = is.na(perp_sex) | is.na(perp_age_group) | is.na(perp_race))
```


# Analysis 1 - Racial Disparities 

At a first glance, one of the most striking features of this dataset that I came across relate to racial disparities. Before we get started, I'll check that all of the racial groups are reasonably well represented in the data with a couple quick cross tabs:

```{r Racial crosstabs}
quick_crosstab <- function(var, focal_df = df) {
  focal_df |> 
  count(.data[[var]]) |> 
  mutate(
    percent = n/sum(n), 
    percent = paste0(round(percent * 100, 1), "%")
  ) |> 
  arrange(n) |> 
  kbl() |> 
  kable_classic_2(full_width = FALSE)
}

quick_crosstab("vic_race", df)
quick_crosstab("perp_race", df)


```

Again, we see that the most common race of the perpetrator is unknown. We will revisit this observation in the next section and, for now, focus on observations when complete perpetrator information. Additionally, there are too few "AMERICAN INDIAN/ALASKAN NATIVE" records to do meaningful analysis and comparisons, so I will drop those records. 

```{r Dropping native records}
racial_df <- df |> 
  filter(vic_race != "AMERICAN INDIAN/ALASKAN NATIVE" & perp_race != "AMERICAN INDIAN/ALASKAN NATIVE")

racial_known_df <- racial_df |> 
  filter(!is.na(perp_race))
```

Let's take another look at the perpetrator cross tab with the missing perp information filtered out: 

```{r Known racial crosstabs}
quick_crosstab("perp_race", racial_known_df)
```

```{r}
# Calculate the date range for the dataset of the statement below
min_year <- df$occur_date |> 
  year() |> 
  min()

max_year <- df$occur_date |> 
  year() |> 
  max()
```


A quick consultation with the [Wikipedia page Race and Ethnicity in New York City](https://en.wikipedia.org/wiki/Race_and_ethnicity_in_New_York_City) suggests that the demographics of shooting victims and perpetrators is inconsistent with that of NYC as a whole. Most dramatically, during the time period of our dataset (`r str_glue("{min_year} to {max_year}")`), we can approximate from the census data reported on the Wikipedia page that 1 in 4 NYC residents is black. However, our cross tab below shows that 7 out of 10 shooting victims is black, and 2 out of 3 known shooting perpetrators is black. 

**Bias Observation #2** As stated above, we have uncovered a prominent racial bias in these shooting incident data. Assuming these data are representative, this amounts to both a statistical bias and a institutional racial bias. However, I want to be careful and acknowledge my personal biases in this case, the complexity of the data, and reframe my research question in light of this. 

1. With these kind of data, it is important to reiterate the adage "correlation is not causation." To understand the drivers of violent crime, we can't look at a single variable in isolation, even when we spot an eye catching pattern. Race and crime in the US is a complex and controversial topic. I do not feel that I have the expertise to expound on the topic. However, [this Wikipedia article on US race and crime](https://en.wikipedia.org/wiki/Race_and_crime_in_the_United_States) is a good introduction to the topic for those interested. In short, for this dataset, if we want to understand the drivers of shooting incidents, we only have a handful of variables (e.g., victim and perp race), but academic research has indicated that a plethora of factors may drive this violence: socioeconomic class, education, housing segregation, and even lead pollution. Most of these are not present in our dataset.
2. Although I want to be measured in my approach to this analysis, I recognize that I personally believe that systemic racism is an important factor in violent crime and policing in the US. This is likely why I choose to look at `vic_race` and `perp_race` as the first analysis of my project.
3. In light of #1 and #2 above, I think it is appropriate to narrow my analysis and ask a more specific question that the data are appropriate for: **given the overrepresentation of certain demographic groups in our data, is these an association between the races of known perpetrators and victims? In other words, are shooting incidents more inter- (between races) or intraracial (within a race) than we might expect?**

Given this new focus, let's visualize the joint distribution of our victim and prepetrator data demographic data with a heatmap: 

```{r ggplot theme}
# I will reuse this theme throughout this report for graphs
cbw_theme <- theme_bw() +  # cbw = centered black and white theme
  theme(plot.title = element_text(hjust = 0.5))

cbw_rotated_theme <- function(angle = -45, hjust = 0, vjust = 1) { # rotate x-axis text 
  cbw_theme + 
  theme(axis.text.x = element_text(angle = angle, hjust = hjust, vjust = vjust))
  }
```

```{r Racial joint distribution plot}

racial_known_df |> 
  filter(!is.na(perp_race)) |> 
  group_by(vic_race, perp_race) |> 
  summarize(cases = n(), .groups = "drop") |> 
  mutate(percent_cases = cases/sum(cases)) |> 
  ggplot(aes(x = vic_race, y = perp_race, fill = percent_cases)) + 
  geom_tile() + 
  cbw_rotated_theme(angle = -25) + 
  labs(
    x = "Victim Race", 
    y = "Perp Race (when known)", 
    title = "Victim-Perp Racial Heatmap", 
    fill = "Percent Incidents"
  ) +
  scale_fill_viridis_c(
    label = label_percent(), 
    option = "magma", 
    guide = guide_colorbar( # Make the color bar less squished
      barwidth = unit(15, "lines"), 
      barheight = unit("0.5", "lines")
    )
  ) +
  theme(
    legend.position = "bottom", 
    legend.justification = "left",
    plot.margin = margin( # Add to right margin to account for rotated x-axis text
      t = 20, b = 20, l = 20, r = 40, 
      unit = "pt"
    ) 
  )

```

Here, the black-black combination stands out as accounting for more than half of incidents, suggesting a lack of statistical independence between victim and perpetrator race. However, we need to remember that this group is overrepresented in the data in general, so we should formally test this with a chi-squared model. 

```{r chi squared}
chi_model <- chisq.test(racial_known_df$vic_race, racial_known_df$perp_race)
print(chi_model)
```

This p-value is highly significant, indicating a statistical association between victim and perpetrator race. We can visualize the model's Pearson residuals to understand which groups influenced this result the most. These residuals follow the formula: (observed - expected) / sqrt(expected). The residuals are, thus, standardized, with positive values indicating that more incidents of a particular racial combination occurred than expected while negative values suggest fewer incidents than expected occurred.

```{r chi squared observed vs. expected}
chi_resid_df <- chi_model$residuals |> 
  as.data.frame() |> 
  rename(
    vic_race = "racial_known_df.vic_race", 
    perp_race = "racial_known_df.perp_race", 
    st_resid = "Freq"
  )

chi_resid_df |> 
  mutate(type = ifelse(vic_race == perp_race, "Intraracial", "Interracial")) |> 
  mutate(vic_race = paste(vic_race, "VICTIMS")) |> 
  ggplot(aes(x = perp_race, y = st_resid, color = type)) + 
    geom_point() + 
    facet_wrap(~vic_race, ncol = 2) + 
    cbw_rotated_theme() + 
    labs(
      title = "Observed vs. Expected Victim and Perp Combinations by Race", 
      x = "Perp Race", 
      y = "Pearson Residuals", 
      color = ""
    ) + 
  scale_color_manual(values = c("Interracial" = "darkblue", "Intraracial" = "darkred"))
```

One pattern that arises from these residuals is that a larger number than expected of intraracial (the victim and perpetrator belong to the same demographic group) incidents is most responsible for the statistical dependence between victim and perpetrator race. Furthermore, the largest residuals come from the unexpectedly high rates of intraracial shootings involving asian/Pacific islander and white groups. I was somewhat surprised by this result since the rate of intraracial shootings is not particularly pronounced within these two groups:

```{r intra percent plots}
racial_known_df |> 
  group_by(vic_race) |> 
  summarize(intraracial_percent = mean(vic_race == perp_race), .groups = "drop") |> 
  mutate(vic_race = fct_reorder(vic_race, intraracial_percent, .desc = TRUE)) |> 
  ggplot(aes(x = vic_race, y = intraracial_percent)) + 
    geom_col(fill = "darkblue") + 
    cbw_rotated_theme() +
    labs(
      x = "Victim Race", 
      y = "Intraracial Percent", 
      title = "Intraracial Shooting Incidents by Victim Race"
    ) + 
  theme(plot.margin = margin(
    t = 20, b = 20, l = 20, r = 40,
    unit = "pt"
  )) + 
  scale_y_continuous(
    labels = label_percent(), 
    breaks = pretty_breaks(n = 10)
  )
```

Overall, we have found a statistical association between the race of victims and perpetrators driven by higher than expected rates of intraracial shootings. This analysis reminds me of the importance of having specific statistical hypotheses and interrogating initial results and visualizations. 

# Analysis 2 - The Hidden World of Perps

The name of the `naniar` package is a reference to C.S. Lewis' Narnia fantasy book series. The package's author sees missing data like the mythical realm of Narnia, a hidden other dimension to our data ([source](https://cran.r-project.org/web/packages/naniar/readme/README.html)). In many statistical analyses, there is a tendency to immediately drop records with missing data and then proceed with further analyses. This can be a critical mistake depending on the type of missing data we are dealing with.

Broadly, there are three types of missing data: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). MCAR is the easiest to deal with since it means that our data are really just a random sample of a theoretical, more complete dataset. MAR is a bit tricker and refers to data that are missing due to the another observed variable in our dataset. For example, in weather data, we might be missing humidity measurements only at high temperatures (also recorded in our dataset) due to the equipment failing in high heat environments. We can correct for MAR with some statistical care. MNAR is the worst case scenario: the missing data are caused by something not captured in our dataset. This is not possible to fix without adding more data that explain the problematic pattern in the missing records. 


```{r Spatial missing data}
df |> 
  filter(!is.na(longitude)) |> 
  mutate(missing_perp_info = ifelse(missing_perp_info, "Missing Perp Info", "Not Missing Perp Info")) |> 
  ggplot(aes(x = longitude, y = latitude, color = missing_perp_info)) + 
  geom_point(alpha = 0.5, shape = 1) + 
  cbw_theme + 
  labs(
    title = "Map of Missing Perp Information", 
    x = "Longitude", 
    y = "Latitude", 
    color = ""
  ) + 
  scale_color_manual(values = c("Not Missing Perp Info" = "#35b701", "Missing Perp Info" = "#9f2b00")) + 
  theme(legend.position = "bottom")
```

Staten Island jumps out as... 

```{r Missing perp by Borro}
df |> 
  group_by(boro) |> 
  summarize(mean_missing_perp_info = mean(missing_perp_info)) |> 
  ungroup() |> 
  mutate(boro = fct_reorder(boro, mean_missing_perp_info, .desc = TRUE)) |> 
  ggplot(aes(x = boro, y = mean_missing_perp_info)) + 
  geom_col(fill = "darkblue") + 
  labs(
    x = "Boro", 
    y = "Percent Records Missing Perp Info", 
    title = "Missing Perp Info by Boro"
  ) + 
  scale_y_continuous(labels = label_percent(), breaks = pretty_breaks(n = 10)) + 
  cbw_rotated_theme()
```

These are fairly substantial differences between the Boros. To make sure this isn't driven by a handful of outliers, let's check the same style plot but with precincts. I will color the bars by Boro: 

```{r Missing perp by precinct}
df |> 
  group_by(precinct, boro) |> 
  summarize(mean_missing_perp_info = mean(missing_perp_info), .groups = "drop") |> 
  mutate(precinct = fct_reorder(precinct, mean_missing_perp_info, .desc = TRUE)) |> 
  ggplot(aes(x = precinct, y = mean_missing_perp_info, fill = boro)) + 
  geom_col() + 
  cbw_theme +
  theme(
    axis.text.x = element_blank(), # Too many precincts to show labels
    axis.ticks.x = element_blank(),
    legend.position = "bottom"
  ) + 
  labs(
    x = "Precinct", 
    y = "Percent Records Missing Perp Info", 
    title = "Missing Perp Info by Precinct", 
    fill = ""
  ) + 
  scale_y_continuous(
    labels = label_percent(),
    breaks = pretty_breaks(n = 10)
  )
```

Since we see a fairly consistent clumping of colors, I will continue using a Boro-level analysis.

```{r Boro models}
# Set reference levels for our categorical variables
mod_df <- df |> 
  mutate(
    vic_race = relevel(as.factor(vic_race), ref = "WHITE"), 
    boro = relevel(as.factor(boro), ref = "MANHATTAN")
  )

boro_race_glm <- glm(missing_perp_info ~ boro + vic_race, data = df, family = "binomial")
summary(boro_race_glm)
```


```{r Missing perp by victim}
df |> 
  group_by(vic_race) |> 
  summarize(mean_missing_perp_info = mean(missing_perp_info)) |> 
  ggplot(aes(x = vic_race, y = mean_missing_perp_info)) + 
  geom_col()
```


```{r upset plot, warning=false}
gg_miss_upset(df) # TODO 260321_1139 Explain this 
```

# Session Info
Please see below for information about the specific versions of the packages used to create this report. 

```{r Session info}
sessionInfo()
```

