---
title: NYPD Shooting Incident Data Report
author: Trey Hendrix
date: |
  Date Created: 260307
  Date Modified: 260323
output: 
  html_document:
    toc: true
    number_sections: true
    
---

```{r Setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE, 
  warning = FALSE
)
```

# Introduction

This report analyzes the...

```{r Packages}
library(tidyverse)
library(kableExtra) # for more aesthetic and customizable data tables
library(jsonlite) # for parsing JSON metadata 
library(naniar) # for visualizing and analyzing missing data
library(scales) # for graph labels
```

```{r R version check, include=FALSE}
# Check for R version 4.1.0 or higher for the base pipe |>
if (getRversion() < "4.1.0") {
  stop(paste(
    "Error: R version 4.1.0 or higher is required to use the base pipe (|>).",
    "Your current version is", getRversion(),
    ". To run this report, please update R at https://cran.r-project.org/"
  ))
}
```

# Importing Data and Metadata

Using the ["ARCHIVED_NYPD Shooting Incident Data (Historic)" on the data.gov website](https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic), we can copy the link for downloading the .csv format of the dataset. Similarly, we can also grab the metadata via this page (which is a JSON).
```{r Import data}
raw_df <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
meta_list <- fromJSON("https://catalog.data.gov/harvest/object/e150b92a-5c72-4adc-8df5-f27a0a0b641b")
```

```{r Data overview}
glimpse(raw_df)
```
Most of these columns are self-explanatory; however, we will need to dig into the metadata to make sure we understand a few of them. 

```{r Column Definitions}
column_meta_list <- meta_list$distribution |> 
  filter(mediaType == "application/json") |> 
  pull(downloadURL) |> 
  fromJSON()

column_def_df <- column_meta_list$meta$view$columns

column_def_df |> 
  filter(fieldName %in% str_to_lower(c("X_COORD_CD", "Y_COORD_CD"))) |> 
  select(fieldName, description)
```

For our geographic data, we will stick to the standard latitude and longitude fields, so I will drop these variables in the section below.

# Data Cleaning

First, I will also perform some quick cleaning, such as consistently coding missing values and renaming columns. 

```{r Variable selection and name cleanup}
null_str_values <- c("(null)", "UNKNOWN") # We will replace these with NAs in R to better identify and deal with missing data

df <- raw_df |> 
  rename_with(.fn = str_to_lower) |>  # make the column names lowercase for consistency
  mutate(
    occur_date = mdy(occur_date), # mdy = month day year date format 
    across( # clean up NA values stored as strings 
      where(is.character), 
      .fn = function(x) {ifelse(x %in% null_str_values, NA_character_, x)}
      ), 
    precinct = as.character(precinct) # These are not meaningful as integers (they are categories)
  ) |> 
  select(-c(x_coord_cd, y_coord_cd)) # drop unwanted columns

```


A few of our variable might also be better stored as ordinal values (i.e., factors in R) since they have a natural ordering. I take care of that below.

```{r Recoding factors}
# TODO 260321_1119 Return and do this 
```

## Missing Data

From the previous `glimpse` call above, it is clear that many records have missing values. The `naniar` R package is my favorite way to understand patterns in missing data. First, let's see which variables have missing records:

```{r Missing counts}
gg_miss_var(df)
miss_var_summary(df)
```

Fortunately, the majority of records only have a small portion of missing records for most variables (`longitude` to `boro` in the graph above). These variables are also fairly essential/fundamental to understanding a crime that I feel comfortable dropping any records that are missing these information. However, a handful of variables are frequently missing. The location information `loc_classfctn_desc` and `loc_of_occur_desc` are missing too frequently to be informative, so I will largely ignore them. `location_desc` is available for 40% of records while information about the perpetrator (`perp_age_group`, `perp_race`, and `perp_sex`) are available for the majority of incidents. I am interested in digging into this pattern of missing data and will do so in [the first analysis section](# Analysis 1 - The Shadow World of Perps), so I will not drop or impute these data.


**Bias Observation #1**: Approximately half of our records are missing information about the race and sex of the perpetrator, while the victims' race and sex is never missing. We need to keep this bias in the data in mind when analyzing perpetrator-victim dynamics. This bias is somewhat intuitive: crimes are reported by victims, not perpetrators.

```{r Critical NA drop}
# Dropping records with critical missing information
df <- df |> 
  drop_na(jurisdiction_code, vic_age_group, vic_race, lon_lat, longitude, latitude)

# Creating a column to track when information is missing about a perp
df <- df |> 
  mutate(missing_perp_info = is.na(perp_sex) | is.na(perp_age_group) | is.na(perp_race))
```


# Analysis 2 - Racial Disparities 

At a first glance, one of the most striking features of this dataset are that... 

Before we get started, I'll check that all of the racial groups are reasonablly well represented in the data with a couple quick crosstabs:
```{r}
quick_crosstab <- function(var) {
  df |> 
  count(.data[[var]]) |> 
  mutate(
    percent = n/sum(n), 
    percent = paste0(round(percent * 100, 1), "%")
  ) |> 
  arrange(n) |> 
  kbl() |> 
  kable_classic_2(full_width = FALSE)
}

quick_crosstab("vic_race")
quick_crosstab("perp_race")


```

Again, we see that the most common race of the perpetrator is unknown... Additionally, there are too few "AMERICAN INDIAN/ALASKAN NATIVE" records to do meaningful analysis and comparisons, so I will drop those records. 

```{r Dropping native records}
racial_df <- df |> 
  filter(vic_race != "AMERICAN INDIAN/ALASKAN NATIVE" & perp_race != "AMERICAN INDIAN/ALASKAN NATIVE")
```


```{r ggplot theme}
# I will reuse this theme throughout this report for graphs
cbw_theme <- theme_bw() +  # cbw = centered black and white theme
  theme(plot.title = element_text(hjust = 0.5))

cbw_rotated_theme <- function(angle = -45, hjust = 0, vjust = 1) { # rotate x-axis text 
  cbw_theme + 
  theme(axis.text.x = element_text(angle = angle, hjust = hjust, vjust = vjust))
  }
```

```{r Racial joint distribution plot}

racial_df |> 
  filter(!is.na(perp_race)) |> 
  group_by(vic_race, perp_race) |> 
  summarize(cases = n(), .groups = "drop") |> 
  mutate(percent_cases = cases/sum(cases)) |> 
  ggplot(aes(x = vic_race, y = perp_race, fill = percent_cases)) + 
  geom_tile() + 
  cbw_rotated_theme(angle = -25) + 
  labs(
    x = "Victim Race", 
    y = "Perp Race (when known)", 
    title = "Victim-Perp Racial Heatmap", 
    fill = "Percent Incidents"
  ) +
  scale_fill_viridis_c(
    label = label_percent(), 
    option = "magma", 
    guide = guide_colorbar( # Make the color bar less squished
      barwidth = unit(20, "lines"), 
      barheight = unit("0.5", "lines")
    )
  ) +
  theme(legend.position = "bottom", legend.justification = "left")

```

To contextualize this dramatic result, I will use 2020 US census data to understand how racial makeup of NYC. The data are available via the [census.gov](https://data.census.gov/table/DECENNIALCD1182020.P9?t=Race+and+Ethnicity&g=500XX00US3605,3606,3607,3608,3609,3610,3611,3612,3613,3614,3615&y=2020) webpage. I am able to approximate the NYC area by selecting congressional districts 5-15 [source](https://en.wikipedia.org/wiki/New_York%27s_congressional_districts). Conveniently, census.gov provides an API URL to access these data, which I have done below. 

```{r Census data}
census_api_url <- "https://api.census.gov/data/2020/dec/cd118?get=group(P9)&ucgid=5001800US3605,5001800US3606,5001800US3607,5001800US3608,5001800US3609,5001800US3610,5001800US3611,5001800US3612,5001800US3613,5001800US3614,5001800US3615"
census_raw_df <- read_csv(census_api_url)
census_raw_df

census_meta_list <- fromJSON("https://api.census.gov/data/2020/dec/cd118") # TODO This is a JSON... decide whether this is worth it to clean these data
# Consider using the tidycensus package too... 
census_vars_list <- fromJSON(census_meta_list$dataset$c_variablesLink)
census_vars_list$variables <- census_vars_list$variables[!names(census_vars_list$variables) %in% c("for", "in", "ucgid")]

census_vars_df <- as_tibble(census_vars_list$variables)

# New idea: use NYC open data xlsx file URL: https://www1.nyc.gov/assets/planning/download/office/data-maps/nyc-population/acs/demo_2016acs5yr_nyc.xlsx
# General URL https://data.cityofnewyork.us/City-Government/Demographic-and-Housing-Profiles-by-Borough/cu9u-3r5e/about_data
```



# Analysis 2 - The Shadow World of Perps

The name of the `naniar` package is a reference to C.S. Lewis' Narnia fantasy book series. The package's author sees missing data like the mythical realm of Narnia, a hidden other dimension to our data ([source](https://cran.r-project.org/web/packages/naniar/readme/README.html))...

Broadly, there are three types of missing data: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). MCAR is the easiest to deal with since it means that our data are really just a random sample of a theoretical, more complete dataset. MAR is a bit tricker and refers to data that are missing due to the another observed variable in our dataset. For example, in weather data, we might be missing humidity measurements only at high temperatures (also recorded in our dataset) due to the equipment failing in high heat environments. We can correct for MAR with some statistical care. MNAR is the worst case scenario: the missing data are caused by something not captured in our dataset. This is not possible to fix without adding more data that explain the problematic pattern in the missing records. 


```{r Spatial missing data}
df |> 
  filter(!is.na(longitude)) |> 
  mutate(missing_perp_info = ifelse(missing_perp_info, "Missing Perp Info", "Not Missing Perp Info")) |> 
  ggplot(aes(x = longitude, y = latitude, color = missing_perp_info)) + 
  geom_point(alpha = 0.5, shape = 1) + 
  cbw_theme + 
  labs(
    title = "Map of Missing Perp Information", 
    x = "Longitude", 
    y = "Latitude", 
    color = ""
  ) + 
  scale_color_manual(values = c("Not Missing Perp Info" = "#35b701", "Missing Perp Info" = "#9f2b00")) + 
  theme(legend.position = "bottom")
```

Staten Island jumps out as... 

```{r Missing perp by Borro}
df |> 
  group_by(boro) |> 
  summarize(mean_missing_perp_info = mean(missing_perp_info)) |> 
  ungroup() |> 
  mutate(boro = fct_reorder(boro, mean_missing_perp_info, .desc = TRUE)) |> 
  ggplot(aes(x = boro, y = mean_missing_perp_info)) + 
  geom_col(fill = "darkblue") + 
  labs(
    x = "Boro", 
    y = "Percent Records Missing Perp Info", 
    title = "Missing Perp Info by Boro"
  ) + 
  scale_y_continuous(labels = label_percent(), breaks = pretty_breaks(n = 10)) + 
  cbw_rotated_theme()
```

These are fairly substantial differences between the Boros. To make sure this isn't driven by a handful of outliers, let's check the same style plot but with precincts. I will color the bars by Boro: 

```{r Missing perp by precinct}
df |> 
  group_by(precinct, boro) |> 
  summarize(mean_missing_perp_info = mean(missing_perp_info), .groups = "drop") |> 
  mutate(precinct = fct_reorder(precinct, mean_missing_perp_info, .desc = TRUE)) |> 
  ggplot(aes(x = precinct, y = mean_missing_perp_info, fill = boro)) + 
  geom_col() + 
  cbw_theme +
  theme(
    axis.text.x = element_blank(), # Too many precincts to show labels
    axis.ticks.x = element_blank(),
    legend.position = "bottom"
  ) + 
  labs(
    x = "Precinct", 
    y = "Percent Records Missing Perp Info", 
    title = "Missing Perp Info by Precinct", 
    fill = ""
  ) + 
  scale_y_continuous(
    labels = label_percent(),
    breaks = pretty_breaks(n = 10)
  )
```

Since we see a fairly consistent clumping of colors, I will continue using a Boro-level analysis.

```{r Boro models}
# Set reference levels for our categorical variables
mod_df <- df |> 
  mutate(
    vic_race = relevel(as.factor(vic_race), ref = "WHITE"), 
    boro = relevel(as.factor(boro), ref = "MANHATTAN")
  )

boro_race_glm <- glm(missing_perp_info ~ boro + vic_race, data = df, family = "binomial")
summary(boro_race_glm)
```


```{r Missing perp by victim}
df |> 
  group_by(vic_race) |> 
  summarize(mean_missing_perp_info = mean(missing_perp_info)) |> 
  ggplot(aes(x = vic_race, y = mean_missing_perp_info)) + 
  geom_col()
```


```{r upset plot, warning=false}
gg_miss_upset(df) # TODO 260321_1139 Explain this 
```

# Session Info
Please see below for information about the specific versions of the packages used to create this report. 

```{r Session info}
sessionInfo()
```

