---
title: NYPD Shooting Incident Data Report
author: Trey Hendrix
date: |
  Date Created: 260307
  Date Modified: 260312
output: 
  html_document:
    toc: true
    number_sections: true
    
---

```{r Setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE
)
```

This report analyzes the...

```{r Packages}
library(tidyverse)
library(kableExtra) # for more aesthetic and customizable data tables
library(jsonlite) # for parsing JSON metadata 
```

```{r R version check, include=FALSE}
# Check for R version 4.1.0 or higher for the base pipe |>
if (getRversion() < "4.1.0") {
  stop(paste(
    "Error: R version 4.1.0 or higher is required to use the base pipe (|>).",
    "Your current version is", getRversion(),
    ". To run this report, please update R at https://cran.r-project.org/"
  ))
}
```

# Importing and Cleaning Data

Using the ["ARCHIVED_NYPD Shooting Incident Data (Historic)" on the data.gov website](https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic), we can copy the link for downloading the .csv format of the dataset. Similarly, we can also grab the metadata via this page (which is a JSON).
```{r Import data}
raw_df <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
meta_list <- fromJSON("https://catalog.data.gov/harvest/object/e150b92a-5c72-4adc-8df5-f27a0a0b641b")
```

```{r Data overview}
glimpse(raw_df)
```
Most of these columns are self-explanatory; however, we will need to dig into the metadata to make sure we understand a few of them. 

```{r Column Definitions}
column_meta_list <- meta_list$distribution |> 
  filter(mediaType == "application/json") |> 
  pull(downloadURL) |> 
  fromJSON()

column_def_df <- column_meta_list$meta$view$columns

column_def_df |> 
  filter(fieldName %in% str_to_lower(c("X_COORD_CD", "Y_COORD_CD"))) |> 
  select(fieldName, description)
```


From the quick overview of the data above, 

# Session Info
Please see below for information about the specific versions of the packages used to create this report. 

```{r Session info}
sessionInfo()
```

